---
title: "COMPETITION ECONOMICS RESEARCH"
author: "Precious Nhamo"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# COMPETITION-ECONOMICS-RESEARCH

This repository contains code for analysing the evolution of South Africa's merger notification thresholds from 1999 to 2026. Using Bloomberg Mergers & Acquisitions (M&A) data, together with CPI, nominal and real GDP, JSE market capitalisation, and the FTSE/JSE All Share Index (ALSI) as benchmarks, we compare actual threshold paths to counterfactual scenarios against other countries.

# DATA

When validating our data do we look for proportions of intermediate to large or do we check how it matches the Commission's Data Set.
<img width="1313" height="590" alt="image" src="https://github.com/user-attachments/assets/ad2cdf29-9b28-4597-abf4-758a4e0538d5" />

# BLOOMBERG DATA

With the bloomberg data I want to classify transactions that would meet the thresholds based on turnover alone, then based on assets alone, and those that would have based on the combined value. When comparing this to the macro variables we may need to use rolling averages, logging the data and structural breaks [check OECD paper for inspiration, they used 3-point rolling average].

What values does our data take and how often?

<img width="1634" height="2105" alt="image" src="https://github.com/user-attachments/assets/f32c030a-5ebf-42fb-a5ba-cc873ce5ed1d" />
<img width="2084" height="1781" alt="image" src="https://github.com/user-attachments/assets/9a45ef57-ee21-4443-8036-85f9bb2501a8" />
<img width="2382" height="1931" alt="image" src="https://github.com/user-attachments/assets/a30c716c-ebe0-49b9-abb0-6bf8e6576117" />
<img width="1933" height="882" alt="image" src="https://github.com/user-attachments/assets/716ad680-4a29-4dac-a505-bf26989f0c47" />
<img width="2233" height="1780" alt="image" src="https://github.com/user-attachments/assets/4c69eb05-af92-40a6-97cc-802bdb4dcb6f" />
<img width="4160" height="1984" alt="sa_merger_notifications" src="https://github.com/user-attachments/assets/af9fd4ff-d9a0-49cd-98d5-ed05577446aa" />
<img width="1935" height="1343" alt="image" src="https://github.com/user-attachments/assets/c0614390-5aa0-418d-9b54-b096a6baa745" />

# MACROECONOMIC VARIABLES

Explain why we use Nominal GDP (its issues with conflating economic activity and price effects in representing changes in national accounts), how this links to sales and inventory and M&A activity in the economy, how real GDP solves this or the GDP deflator to account for local nexus and market capitalisation, is GNP an alternative and GDP measured in PPP and what is the economic rationale for choosing that? Could we have used other variables? And lastly market capitalisation and the ALSI.

<img width="1484" height="1331" alt="image" src="https://github.com/user-attachments/assets/6b1c76e3-14cc-4a5b-ba36-46b6ed32edee" />
<img width="2385" height="1916" alt="image" src="https://github.com/user-attachments/assets/d7d242ad-a572-4881-8e94-da04c65fa902" />
<img width="4160" height="1984" alt="image" src="https://github.com/user-attachments/assets/3b2cec42-cfb8-4174-b825-186bf29700c7" />

## Why both JSE market capitalisation and the ALSI are retained

Full JSE market capitalisation is the nominal rand value of the whole market and is the variable that most closely matches the Competition Commission's own historical wording ("market capitalisation on the Johannesburg Stock Exchange"). The FTSE/JSE All Share Index (ALSI, J203) is instead a price/capital index covering roughly 99% of eligible Main Board securities by value — an index level, not a rand value — and it is **not a substitute for market capitalisation**: over 2017→2026 the two benchmarks diverge sharply (ALSI +91.0% to a 19 May 2026 snapshot vs +59.9% for full market cap to a 5 June 2026 primary-source snapshot). Both are therefore retained as separate benchmarks: market capitalisation as the primary market-based benchmark (matching the Commission's own language), and the ALSI as a robustness/equity-price comparison. Neither series has a completed full 2026 annual observation; both 2026 points below are dated, point-in-time snapshots, not annual averages.

*Standard workflow reminder, following initial inspection (structure, missingness, summary stats, distribution plots): 1) cleaning decisions, made explicit; 2) feature engineering; 3) bivariate/relationship exploration; 4) formal statistical tests where relevant; 5) model specification and estimation; 6) diagnostics and robustness; 7) communicating results.*

## Construction of the Transaction Sample

**Source.** We evaluate four Bloomberg MA<GO> exports totalling 13,483 rows: two queries, split at 20 August 2015, each run twice with different columns. Matching on type, date, target and acquirer, the ticker exports strictly contain the others, which lose 710 deals and add none; shared values agree throughout. We take the ticker exports as the row universe, 7,105 rows.

Bloomberg data are compiled from filings, press releases, news wires and direct submissions, and unlisted firms are covered subject to disclosure, and coverage "can be thinner for smaller or non-reportable transactions." *Transactions near the intermediate threshold are therefore under-represented, so counts of deals sitting just below the notification boundary are lower bounds.* Sector fields return current, not point-in-time, classifications. *A firm acquired in 2003 carries its 2026 sector, so sector results describe classification today rather than at the deal date.*

**Defects.** Sixteen rows carry a negative turnover across only three distinct values and three firms; we read this as a sign error and correct it. *Left negative, these firms would fail the turnover route automatically and be recorded as asset-driven, so the correction changes which measure is decisive for them.* One deal value is overstated a thousandfold, at forty-eight times the acquirer's balance sheet, which we flag and do not use. *Deal value enters no statutory test, so this affects descriptive statistics on transaction size only.*

The financial columns are firm-level: of 238 acquirers appearing three or more times, every one carries an identical assets figure across all its deals. *These are each firm's latest accounts, not its position at announcement. The median transaction predates the export by fifteen years, so early deals are tested at present-day size against thresholds set in their own era, overstating how many cleared them.* And 11 per cent of targets are asset descriptions — mineral blocks, property portfolios, tower sites. *Missingness is concentrated on the target side by construction: 53 of the 70 three-value records lack a target-side figure, so transferred-firm results rest on a smaller, non-random subset.*

**Selection.** Section 12(1)(b) provides that a merger may be achieved through purchase of shares, an interest, or assets, so we retain deal types M&A, INV and AST. *Excluding asset acquisitions would have narrowed the population below the statutory definition.* We remove 133 repurchases and unbundlings, which record Shareholders as the acquirer and involve no acquisition of control, 106 joint ventures, 15 firms acquiring their own shares, and 166 naming no acquirer or a placeholder. *The repurchases are 28 complete records of large listed firms; retaining them would tilt the sample further towards that group. The placeholders carry no complete records, so that exclusion is immaterial.* We collapse repeated target–acquirer pairs announced within 365 days, keeping the fullest record. *This is the only judgement in the pipeline that materially moves the sample size; a revised bid left uncollapsed would double-weight the same firms.*

**Result.** This leaves 6,511 unique transactions, of which 476 carry three or four financial values: 406 complete and 70 with three, announced between February 1999 and July 2026. Thirty are Withdrawn or Proposed, which Bloomberg defines as rumoured or non-binding, with no definitive agreement signed; all are complete records and all post-date 2009. We retain them, identifiable by status. *They are not a random thirty: retaining them tilts the sample towards recent large listed transactions. Dropping them gives 446.*

**Cleaning funnel**

| Step | Rows |
|---|---:|
| Raw rows across the four exports | 13,483 |
| Ticker exports only | 7,105 |
| After exact and boundary duplicates | 7,095 |
| Deal types M&A, INV and AST | 6,856 |
| Acquirer and target distinct firms | 6,841 |
| Acquirer is a named firm | 6,675 |
| **Unique transactions** | **6,511** |
| &nbsp;&nbsp;with 0 of 4 financial values | 2,362 |
| &nbsp;&nbsp;with 1 of 4 | 398 |
| &nbsp;&nbsp;with 2 of 4 | 3,275 |
| &nbsp;&nbsp;with 3 of 4 | 70 |
| &nbsp;&nbsp;with 4 of 4 | 406 |
| **ANALYSIS SAMPLE (3 or 4 values)** | **476** |
| &nbsp;&nbsp;of which Withdrawn or Proposed | 30 |

*Source: Bloomberg MA<GO>.*

# Data Analysis

```{r load-data, echo=FALSE, message=FALSE}
# I don't want all of the information to be printed out here
library(readr)
library(dplyr)
madata <- read_csv("data/ma_classification_blended.csv", show_col_types = FALSE)
str(madata)
dim(madata)
```

## Adding JSE market capitalisation and the ALSI to the blended data

The blended dataset already carries full JSE market capitalisation, sourced consistently with the dedicated JSE/ALSI workbook (cross-checked below to within 0.04% at every threshold-revision year). The ALSI series is new and is merged in here by year.

```{r merge-alsi, message=FALSE}
library(readxl)
library(tidyr)

jse_file <- "data/JSE_market_cap_and_ALSI_1999_2026.xlsx"

alsi_annual <- read_excel(jse_file, sheet = "ALSI_Annual", skip = 3) %>%
  select(Year, ALSI = `ALSI level`)

mcap_annual <- read_excel(jse_file, sheet = "JSE_Market_Cap", skip = 3) %>%
  select(Year, `Market cap (R million)`)

# 2026 has two snapshot rows in each sheet (dated point-in-time observations,
# not a completed annual figure). Keep the later/primary one in each case:
# ALSI -> 29 May 2026 snapshot; Market cap -> 5 June 2026 JSE-primary snapshot.
alsi_annual <- alsi_annual %>%
  filter(Year != 2026) %>%
  bind_rows(alsi_annual %>% filter(Year == 2026) %>% slice(2))

mcap_annual <- mcap_annual %>%
  filter(Year != 2026) %>%
  bind_rows(mcap_annual %>% filter(Year == 2026) %>% slice(2))

# Cross-check the new file's market cap against what's already in madata
# (should agree to within rounding -- confirms the two sources are consistent)
madata %>%
  distinct(Year, .keep_all = TRUE) %>%
  select(Year, `Market Capitalisation (R million)`) %>%
  inner_join(mcap_annual, by = "Year") %>%
  filter(Year %in% c(2001, 2009, 2017)) %>%
  mutate(diff_pct = round((`Market cap (R million)` - `Market Capitalisation (R million)`) /
                             `Market Capitalisation (R million)` * 100, 3))

# Add ALSI onto madata by Year (market cap already present and verified above)
madata <- madata %>%
  left_join(alsi_annual, by = "Year")
```

## Evolution of the SA threshold (growth by policy change)

```{r threshold-growth-chart}
# ============================================================
# Threshold Growth Chart — Small Multiples, Independent Y-Axes
# ============================================================
library(dplyr)
library(tidyr)
library(ggplot2)

thresholds <- tibble(
  Limb      = c("Intermediate: Combined", "Intermediate: Target",
                "Large: Combined", "Large: Target"),
  base_1999 = c(50, 5, 3500, 100),
  v2001     = c(200, 30, 3500, 100),
  v2009     = c(560, 80, 6600, 190),
  v2017     = c(600, 100, 6600, 190),
  v2026     = c(1000, 200, 9500, 280)
)

long <- thresholds %>%
  pivot_longer(-Limb, names_to = "period_end", values_to = "value") %>%
  group_by(Limb) %>%
  mutate(
    increment  = value - lag(value, default = 0),
    pct_growth = (value / lag(value) - 1) * 100,
    period = factor(period_end,
                     levels = c("base_1999", "v2001", "v2009", "v2017", "v2026"),
                     labels = c("1999 base", "+2001", "+2009", "+2017", "+2026"))
  ) %>%
  ungroup()

ggplot(long, aes(x = 1, y = increment, fill = period)) +
  geom_col(width = 0.6, color = "white") +
  geom_text(
    aes(label = case_when(
      period == "1999 base" ~ paste0("R", value, "m"),
      increment > 0         ~ paste0("+", round(pct_growth), "%"),
      TRUE                  ~ ""
    )),
    position = position_stack(vjust = 0.5),
    color = "white", fontface = "bold", size = 3
  ) +
  facet_wrap(~ Limb, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = c(
    "1999 base" = "#D9D9D9", "+2001" = "#35588A",
    "+2009" = "#A83232", "+2017" = "#4C8C6B", "+2026" = "#B8892B"
  )) +
  labs(
    y = "Threshold Value (R'm)", x = NULL, fill = NULL,
    title = "Growth of Each Threshold Limb Since 1999",
    subtitle = "Each panel independently scaled"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )

ggsave("threshold_growth_small_multiples.png", width = 16, height = 9, dpi = 150)
```

### Grayscale version (for print)

In nominal terms, South Africa's merger notification thresholds have evolved in uneven and episodic steps rather than through smooth or uniform adjustment. The most pronounced increase occurred in the 1999–2001 interval, when the intermediate combined threshold rose by 300 per cent and the intermediate target threshold by 500 per cent, while the large-merger thresholds remained unchanged. A second major adjustment followed in 2001–2009, when all four thresholds increased, with nominal growth again stronger for intermediate mergers than for large mergers. By contrast, the 2009–2017 interval was characterised by very limited adjustment: the intermediate combined threshold increased by only 7.1 per cent, the intermediate target threshold by 25 per cent, and the large thresholds did not change. The 2017–2026 revision marked a renewed upward adjustment, with the intermediate target threshold doubling, the intermediate combined threshold rising by 66.7 per cent, and the large thresholds increasing by 43.9 per cent and 47.4 per cent respectively. Overall, the nominal pattern indicates that threshold growth has been concentrated in a few discrete regulatory revisions, with particularly strong increases at the intermediate-merger boundary.

```{r threshold-growth-grayscale}
grays <- c("1999 base" = "#E8E8E8", "+2001" = "#BFBFBF", "+2009" = "#8C8C8C",
           "+2017" = "#595959", "+2026" = "#262626")

long <- long %>%
  mutate(label_color = if_else(period %in% c("1999 base", "+2001"), "black", "white"))

ggplot(long, aes(x = 1, y = increment, fill = period)) +
  geom_col(width = 0.6, color = "white", linewidth = 0.4) +
  geom_text(
    aes(label = case_when(
          period == "1999 base" ~ paste0("R", value, "m"),
          increment > 0         ~ paste0("+", round(pct_growth), "%"),
          TRUE                  ~ ""),
        color = label_color),
    position = position_stack(vjust = 0.5), fontface = "bold", size = 3, show.legend = FALSE
  ) +
  scale_color_identity() +
  facet_wrap(~ Limb, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = grays) +
  labs(y = "Threshold Value (R'm)", x = NULL, fill = NULL,
       title = "Growth of Each Threshold Limb Since 1999",
       subtitle = "Grayscale — each panel independently scaled") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "top", strip.text = element_text(face = "bold"))

ggsave("threshold_grayscale.png", width = 16, height = 9, dpi = 150)
```

## Research questions and empirical approach

The empirical analysis asks how South Africa's merger notification thresholds have evolved, whether their adjustment can be rationalised by movements in observable measures of the economy, and how South Africa's approach compares with alternative methods of setting merger thresholds. Six benchmarks are now used throughout: CPI, the GDP deflator, nominal GDP, real GDP, full JSE market capitalisation, and the FTSE/JSE ALSI.

### Question 1: What if the Commission had chosen a macroeconomic benchmark in 2001?

Had the Competition Commission chosen in 2001 to anchor the merger notification thresholds to a macroeconomic benchmark, what would each of the four nominal thresholds have been at the subsequent revision points in 2009, 2017 and 2026? For each benchmark, I calculate the percentage growth between 2001 and each subsequent revision date and use this growth to construct the corresponding counterfactual intermediate combined, intermediate target-firm, large combined and large target-firm thresholds, then compare these counterfactual values with the thresholds actually implemented to determine which benchmark most closely tracks the observed statutory path.

I use 2001 rather than 1999 as the principal starting point. The 1999 values were the initial thresholds under the new merger-control regime, while the 2001 revision followed an early review of merger activity. The 1999–2001 adjustment is therefore reported descriptively but is not treated as evidence of a systematic threshold-adjustment rule. Njisane et al. similarly describe the 2001 change as following a review of merger trends since the inception of the Competition Act.

```{r question-1, message=FALSE}
# ============================================================
# QUESTION 1: "Had the commission made an implicit or explicit
# choice to pin the thresholds down to the macroeconomic
# variables we have in our data in 2001 (given that the 1999
# were their initial threshold series, the revision shows
# initial calibration after lessons, so I wouldn't read too
# much into the 1999-2001 revisions) what would each values
# have been at the revision points & what percentage growth
# would they have stipulated, and with commission's current
# data which macro benchmark predicted that better?"
#
# UPDATED: now six benchmarks (CPI, GDP deflator, nominal GDP,
# real GDP, JSE market cap, ALSI) instead of five.
# ============================================================

library(dplyr)
library(tidyr)
library(purrr)
library(kableExtra)

# ---- Macro series: CPI/GDP/Deflator from madata; market cap + ALSI
# from the merged JSE/ALSI series added above ----
macro <- madata %>%
  distinct(Year, .keep_all = TRUE) %>%
  transmute(Year,
            CPI  = `CPI Index (Dec 2024=100)`,
            NGDP = `Nominal GDP (R million)`,
            RGDP = `Real GDP (R million, 2015 prices)`,
            DEFL = `GDP Deflator (Index, 2015=100)`,
            MCAP = `Market Capitalisation (R million)`,
            ALSI = ALSI) %>%
  arrange(Year) %>%
  filter(!is.na(CPI))  # drops the incomplete 2026 row in madata itself

# ---- 2026-so-far values for the GDP-derived series (partial-year, as of
# Sept 2026 -- CPI: Jan-Jul avg; Real/Nominal GDP: Q1-Q2/Q1 annualised).
# Market cap and ALSI 2026 already came from the dated JSE/ALSI snapshots
# merged in above, so only the GDP-derived figures need overriding here. ----
# ---- 2026-so-far values (partial-year, as of Sept 2026).
# CPI: Jan-Jul avg; Real/Nominal GDP: Q1-Q2/Q1 annualised (GDP-derived,
# entered here). Market cap and ALSI 2026 come from the dated JSE/ALSI
# snapshots already sitting in mcap_annual / alsi_annual from the merge
# step above -- pulled in here rather than hardcoded a second time. ----

row_2026 <- tibble(
  Year = 2026,
  CPI  = 105.8,
  NGDP = 8025000,
  RGDP = 4720000,
  DEFL = 8025000 / 4720000 * 100,
  MCAP = mcap_annual %>% filter(Year == 2026) %>% pull(`Market cap (R million)`),
  ALSI = alsi_annual %>% filter(Year == 2026) %>% pull(ALSI)
)

macro <- macro %>%
  filter(Year != 2026) %>%   # drop any incomplete/partial 2026 row first
  bind_rows(row_2026) %>%
  arrange(Year)

# sanity check
macro %>% filter(Year == 2026)
macro <- macro %>%
  rows_update(
    tibble(Year = 2026, CPI = 105.8, NGDP = 8025000, RGDP = 4720000,
           DEFL = 8025000/4720000*100),
    by = "Year"
  )

# ---- Threshold data (Table 2, legislated figures) ----
actual <- tribble(
  ~Year, ~IC,   ~IT,  ~LC,   ~LT,
  1999,   50,    5,   3500,  100,
  2001,   200,   30,  3500,  100,
  2009,   560,   80,  6600,  190,
  2017,   600,   100, 6600,  190,
  2026,   1000,  200, 9500,  280
)
limb_names  <- c(IC = "Interm: Combined", IT = "Interm: Target",
                  LC = "Large: Combined",  LT = "Large: Target")
bench_names <- c(CPI = "CPI", NGDP = "Nominal GDP", RGDP = "Real GDP",
                  DEFL = "GDP Deflator", MCAP = "Market Cap", ALSI = "ALSI")
base_year <- 2001

predict_years <- c(2009, 2017, 2026)
base_macro  <- macro %>% filter(Year == base_year)
base_thresh <- actual %>% filter(Year == base_year)

q1_data <- map_dfr(predict_years, function(yr) {
  yr_macro <- macro %>% filter(Year == yr)
  map_dfr(names(limb_names), function(code) {
    actual_val <- actual %>% filter(Year == yr) %>% pull(all_of(code))
    base_val   <- base_thresh %>% pull(all_of(code))
    map_dfr(names(bench_names), function(b) {
      gf <- yr_macro[[b]] / base_macro[[b]]
      predicted <- base_val * gf
      tibble(Year = yr, Limb = limb_names[[code]], Benchmark = bench_names[[b]],
             `Predicted (R'm)` = round(predicted, 1),
             `Predicted Growth %` = round((gf - 1) * 100, 1),
             `Error vs Actual %` = round((predicted - actual_val) / actual_val * 100, 1))
    })
  })
})

# ---- ONE INVERTED (metrics-as-rows) TABLE PER YEAR, sized to fit a page ----
for (yr in predict_years) {
  yr_block <- q1_data %>% filter(Year == yr)
  wide <- yr_block %>%
    pivot_longer(c(`Predicted (R'm)`, `Predicted Growth %`, `Error vs Actual %`),
                 names_to = "Metric", values_to = "value") %>%
    unite("row_label", Benchmark, Metric, sep = " \u2014 ") %>%
    pivot_wider(id_cols = row_label, names_from = Limb, values_from = value)

  actual_row <- actual %>% filter(Year == yr) %>%
    pivot_longer(-Year, names_to = "code", values_to = "val") %>%
    mutate(Limb = limb_names[code]) %>%
    select(Limb, val) %>%
    pivot_wider(names_from = Limb, values_from = val) %>%
    mutate(row_label = "Actual Value (R'm)", .before = 1)

  final_tbl <- bind_rows(actual_row, wide)

  print(
    kbl(final_tbl, format = "latex", booktabs = TRUE, longtable = FALSE,
        col.names = c("", names(final_tbl)[-1]),
        caption = paste0("Threshold values predicted under each macro benchmark vs actual, ", yr,
                          " (single base = 2001)"),
        label = paste0("tab:q1-", yr)) %>%
      kable_styling(latex_options = c("scale_down", "hold_position"),
                    font_size = 8) %>%
      column_spec(1, width = "5cm") %>%
      row_spec(0, bold = TRUE)
  )
}
```

### Question 2: What if the benchmark were reconsidered at each revision?

Had the Commission reconsidered the appropriate benchmark at each threshold revision, what adjustment would each macroeconomic benchmark have implied over the individual intervals 2001–2009, 2009–2017 and 2017–2026? Unlike Question 1, the thresholds are re-anchored to the actual statutory value at the beginning of each interval. This allows the analysis to ask whether different revisions appear to have followed different economic benchmarks rather than assuming a single adjustment rule throughout the entire period.

This is particularly relevant because the Commission's stated rationale has differed across revisions. Njisane et al. report that the 2009 adjustment was informed by nominal GDP and market capitalisation, whereas the 2017 adjustment was based on real GDP growth.

```{r question-2, message=FALSE}
# ============================================================
# QUESTION 2: "Had this choice been made within interval, that
# is reanchoring to 2001, 2009, 2017, 2026, what would have
# this implied, given the commission's reports states different
# reasons at each interval? Just verify the macro benchmark
# data because some of them are already indexed to 2024 etc
# (handle that)."
#
# Note: base-year indexing is handled automatically here because
# every comparison uses a ratio of the index at two points in
# time -- the arbitrary base year (Dec 2024 for CPI, 2015 for the
# GDP Deflator) cancels out algebraically in a ratio.
# 1999-2001 is EXCLUDED: it's the initial post-enactment
# calibration, not a benchmark-driven revision (see Question 1
# framing above and the negative-control check in the appendix).
#
# UPDATED: now six benchmarks, ALSI added.
# ============================================================

intervals <- list(c(2001, 2009), c(2009, 2017), c(2017, 2026))

q2_data <- map_dfr(intervals, function(iv) {
  start_yr <- iv[1]; end_yr <- iv[2]
  start_macro <- macro %>% filter(Year == start_yr)
  end_macro   <- macro %>% filter(Year == end_yr)
  start_thresh <- actual %>% filter(Year == start_yr)
  end_thresh   <- actual %>% filter(Year == end_yr)

  map_dfr(names(limb_names), function(code) {
    a_start <- start_thresh %>% pull(all_of(code))
    a_end   <- end_thresh %>% pull(all_of(code))
    map_dfr(names(bench_names), function(b) {
      gf <- end_macro[[b]] / start_macro[[b]]
      implied <- a_start * gf
      tibble(Interval = paste0(start_yr, "\u2192", end_yr), Limb = limb_names[[code]],
             Benchmark = bench_names[[b]],
             `Implied (R'm)` = round(implied, 1),
             `Implied Growth %` = round((gf - 1) * 100, 1),
             `Error vs Actual %` = round((implied - a_end) / a_end * 100, 1))
    })
  })
})

# ---- ONE INVERTED TABLE PER INTERVAL ----
for (iv in intervals) {
  label <- paste0(iv[1], "\u2192", iv[2])
  iv_block <- q2_data %>% filter(Interval == label)
  wide <- iv_block %>%
    pivot_longer(c(`Implied (R'm)`, `Implied Growth %`, `Error vs Actual %`),
                 names_to = "Metric", values_to = "value") %>%
    unite("row_label", Benchmark, Metric, sep = " \u2014 ") %>%
    pivot_wider(id_cols = row_label, names_from = Limb, values_from = value)

  actual_rows <- bind_rows(
    actual %>% filter(Year == iv[1]) %>%
      pivot_longer(-Year, names_to="code", values_to="val") %>%
      mutate(Limb=limb_names[code]) %>% select(Limb,val) %>%
      pivot_wider(names_from=Limb, values_from=val) %>%
      mutate(row_label = paste0("Actual Start (", iv[1], ", R'm)"), .before=1),
    actual %>% filter(Year == iv[2]) %>%
      pivot_longer(-Year, names_to="code", values_to="val") %>%
      mutate(Limb=limb_names[code]) %>% select(Limb,val) %>%
      pivot_wider(names_from=Limb, values_from=val) %>%
      mutate(row_label = paste0("Actual End (", iv[2], ", R'm)"), .before=1)
  )

  final_tbl <- bind_rows(actual_rows, wide)

  print(
    kbl(final_tbl, format = "latex", booktabs = TRUE,
        col.names = c("", names(final_tbl)[-1]),
        caption = paste0("Threshold growth reanchored per interval, ", label,
                          " \u2014 which macro benchmark implied best fit (six benchmarks)"),
        label = paste0("tab:q2-", gsub("\u2192","-",label))) %>%
      kable_styling(latex_options = c("scale_down", "hold_position"),
                    font_size = 8) %>%
      column_spec(1, width = "5cm") %>%
      row_spec(0, bold = TRUE)
  )
}
```

**Updated best-fit summary (six benchmarks).** Adding the ALSI changes two results relative to the five-benchmark version: ALSI is now the closest-fitting benchmark for both Intermediate limbs over 2001→2009 (errors of −5.4% and −0.6%, ahead of Nominal GDP's −14.4%/−10.1%), and for Intermediate: Target over 2017→2026 (−3.7%, ahead of Market Cap's −20.0% on the updated, primary-sourced market-cap figure). Real GDP remains the unique best fit across all four limbs for 2009→2017, and CPI remains the best fit for both Large limbs over 2017→2026 (errors of 2.9% and 0.5%) — those two findings are unchanged by adding the ALSI.

### Question 3: How large were the revisions relative to alternative threshold-setting methods?

How do the actual statutory adjustments compare with the threshold increases examined by Njisane et al. and with the World Bank's economy-size benchmarking approach?

Njisane et al.'s scenarios are used as sensitivity benchmarks. Their target-firm thresholds are increased by 5%, 10%, 15% and 20%, while combined thresholds are increased by 10%, 20%, 30%, 40% and 45%. They also examine joint increases of 15% in the target threshold with either 20% or 40% in the combined threshold. These percentage bands are compared with the actual percentage change in each South African threshold over the 2001–2009, 2009–2017 and 2017–2026 intervals.

As a sensitivity exercise, the percentage-adjustment scenarios considered by Njisane et al. are applied to the statutory threshold in force at the beginning of each revision interval. This extends their scenario-based approach, originally applied to the October 2017 thresholds and the 2016–2018 merger database, to the historical threshold path. How does the magnitude of the actual revision compare with moderate threshold shocks that prior South African research considered plausible enough to test for error-cost consequences?

```{r question-3-njisane, message=FALSE}
library(tidyverse)
library(kableExtra)

# ============================================================
# QUESTION 3 (Njisane et al. 2021 sensitivity comparison)
# How large were SA's actual threshold revisions relative to
# the target/combined increases Njisane et al. tested, and what
# would their two selected joint scenarios have implied at each
# SA revision point?
# ============================================================

# ---- 1. Actual SA threshold history (R million) ----
sa_thresholds <- tribble(
  ~year, ~int_combined, ~int_target, ~large_combined, ~large_target,
  2001,      200,           30,           3500,          100,
  2009,      560,           80,           6600,          190,
  2017,      600,          100,           6600,          190,
  2026,     1000,          200,           9500,          280
)

actual_growth <- sa_thresholds %>%
  pivot_longer(-year, names_to = "limb_code", values_to = "threshold") %>%
  group_by(limb_code) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    start_year        = lag(year),
    start_threshold   = lag(threshold),
    interval          = paste0(start_year, "\u2192", year),
    actual_growth_pct = 100 * (threshold / start_threshold - 1)
  ) %>%
  ungroup() %>%
  filter(!is.na(start_year)) %>%
  mutate(
    limb_type = if_else(str_detect(limb_code, "target"), "Target", "Combined"),
    limb = recode(limb_code,
                  int_combined   = "Intermediate: Combined",
                  int_target     = "Intermediate: Target",
                  large_combined = "Large: Combined",
                  large_target   = "Large: Target")
  ) %>%
  select(interval, limb, limb_type,
         start_threshold, actual_end_threshold = threshold, actual_growth_pct)

# ---- 2. Njisane et al. (2021) single-limb sensitivity scenarios ----
njisane_scenarios <- tribble(
  ~limb_type, ~scenario_pct,
  "Target",    5,  "Target",   10, "Target",   15, "Target",   20,
  "Combined", 10,  "Combined", 20, "Combined", 30, "Combined", 40, "Combined", 45
)

njisane_range <- njisane_scenarios %>%
  group_by(limb_type) %>%
  summarise(njisane_min = min(scenario_pct),
            njisane_max = max(scenario_pct), .groups = "drop")

closest_njisane <- actual_growth %>%
  left_join(njisane_scenarios, by = "limb_type") %>%
  mutate(diff_pp = actual_growth_pct - scenario_pct) %>%
  group_by(interval, limb) %>%
  slice_min(abs(diff_pp), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(interval, limb, closest_njisane_pct = scenario_pct, closest_diff_pp = diff_pp)

# ---- 3. Njisane's two selected JOINT scenarios ----
joint_scenarios <- tribble(
  ~scenario_label,             ~target_pct, ~combined_pct,
  "15%T / 20%C implied",       15,          20,
  "15%T / 40%C implied",       15,          40
)

njisane_joint_wide <- actual_growth %>%
  select(interval, limb, limb_type, start_threshold) %>%
  crossing(joint_scenarios) %>%
  mutate(
    applied_pct       = if_else(limb_type == "Target", target_pct, combined_pct),
    implied_threshold = start_threshold * (1 + applied_pct / 100)
  ) %>%
  select(interval, limb, scenario_label, implied_threshold) %>%
  pivot_wider(names_from = scenario_label, values_from = implied_threshold)

# ---- 4. ONE combined table ----
q3_table <- actual_growth %>%
  left_join(njisane_range,      by = "limb_type") %>%
  mutate(
    njisane_range_label = paste0(njisane_min, "\u2013", njisane_max, "%"),
    position_vs_range = case_when(
      actual_growth_pct < njisane_min ~ "Below range",
      actual_growth_pct > njisane_max ~ "Above range",
      TRUE ~ "Within range"
    )
  ) %>%
  left_join(closest_njisane,    by = c("interval", "limb")) %>%
  left_join(njisane_joint_wide, by = c("interval", "limb")) %>%
  transmute(
    Interval              = interval,
    Limb                  = limb,
    `Start (R'm)`         = round(start_threshold, 1),
    `Actual End (R'm)`    = round(actual_end_threshold, 1),
    `Actual Growth %`     = round(actual_growth_pct, 1),
    `Njisane Range`       = njisane_range_label,
    `Position`            = position_vs_range,
    `Closest Njisane %`   = closest_njisane_pct,
    `Diff (pp)`           = round(closest_diff_pp, 1),
    `15%T/20%C (R'm)`     = round(`15%T / 20%C implied`, 1),
    `15%T/40%C (R'm)`     = round(`15%T / 40%C implied`, 1)
  ) %>%
  arrange(Interval, Limb)

print(
  kbl(q3_table, format = "latex", booktabs = TRUE, longtable = TRUE,
      caption = "SA threshold revisions vs. Njisane et al. (2021) sensitivity scenarios",
      label = "tab:q3-njisane") %>%
    kable_styling(latex_options = c("scale_down", "hold_position", "repeat_header"),
                  font_size = 7) %>%
    row_spec(0, bold = TRUE)
)



# ---- 5. ONE graph: actual growth vs Njisane's tested range, faceted by interval ----
plot_data <- actual_growth %>% left_join(njisane_range, by = "limb_type")

ggplot(plot_data, aes(x = limb, y = actual_growth_pct)) +
  geom_col(fill = "grey40", width = 0.6) +
  geom_errorbar(aes(ymin = njisane_min, ymax = njisane_max),
                width = 0.15, linewidth = 0.7, colour = "black") +
  facet_wrap(~ interval, nrow = 1, scales = "free_x") +
  labs(
    title = "Actual SA threshold growth vs. Njisane et al. (2021) tested range",
    subtitle = "Bars: actual nominal growth. Error bars: min\u2013max of Njisane's tested scenarios for that limb type.",
    x = NULL, y = "Nominal threshold growth (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Macroeconomic data and measurement

All benchmark comparisons are based on growth rates rather than the absolute levels of the macroeconomic indices. Rebasings therefore do not affect the results provided that each growth calculation uses a consistently linked historical series — this applies equally to the CPI (Dec 2024 = 100), the GDP deflator (2015 = 100), and the ALSI (a price index with no fixed base year in this compilation), since every comparison below uses a ratio of the index at two points in time and the arbitrary base cancels out algebraically.

CPI is measured using the official headline CPI series. The GDP deflator is treated as a price-level measure and is constructed or obtained from a consistent national-accounts vintage. Nominal GDP is measured at current prices, while real GDP is measured at constant 2015 prices. Full JSE market capitalisation is the nominal rand value of the whole market; the ALSI is the FTSE/JSE All Share Index (J203), an equity price/capital index rather than a rand value, retained as a distinct robustness benchmark for the reasons set out above.

For each benchmark \(B\), the counterfactual threshold is calculated as:

\[
T_{t}^{B}=T_{s}\left(\frac{B_t}{B_s}\right),
\]

where \(T_s\) is the statutory threshold at the chosen anchor date, \(B_s\) is the benchmark value at that date, and \(B_t\) is its value at the subsequent revision date. The implied percentage adjustment is:

\[
g_{s,t}^{B}=\left(\frac{B_t}{B_s}-1\right)\times100.
\]

For Question 1, \(s=2001\) throughout. For Question 2, \(s\) is reset to the statutory threshold at the beginning of each interval: 2001, 2009 and 2017.
