---
title: "COMPETITION ECONOMICS RESEARCH"
author: "Precious Nhamo"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# COMPETITION-ECONOMICS-RESEARCH

This repository contains code for analysing the evolution of South Africa's merger notification thresholds from 1999 to 2026. Using Bloomberg Mergers & Acquisitions (M&A) data, together with CPI, nominal and real GDP, JSE market capitalisation, and the FTSE/JSE All Share Index (ALSI) as benchmarks, we compare actual threshold paths to counterfactual scenarios against other countries.

# DATA

When validating our data do we look for proportions of intermediate to large or do we check how it matches the Commission's Data Set.
<img width="1313" height="590" alt="image" src="https://github.com/user-attachments/assets/ad2cdf29-9b28-4597-abf4-758a4e0538d5" />

# BLOOMBERG DATA

With the bloomberg data I want to classify transactions that would meet the thresholds based on turnover alone, then based on assets alone, and those that would have based on the combined value. When comparing this to the macro variables we may need to use rolling averages, logging the data and structural breaks [check OECD paper for inspiration, they used 3-point rolling average].

What values does our data take and how often?

<img width="1634" height="2105" alt="image" src="https://github.com/user-attachments/assets/f32c030a-5ebf-42fb-a5ba-cc873ce5ed1d" />
<img width="2084" height="1781" alt="image" src="https://github.com/user-attachments/assets/9a45ef57-ee21-4443-8036-85f9bb2501a8" />
<img width="2382" height="1931" alt="image" src="https://github.com/user-attachments/assets/a30c716c-ebe0-49b9-abb0-6bf8e6576117" />
<img width="1933" height="882" alt="image" src="https://github.com/user-attachments/assets/716ad680-4a29-4dac-a505-bf26989f0c47" />
<img width="2233" height="1780" alt="image" src="https://github.com/user-attachments/assets/4c69eb05-af92-40a6-97cc-802bdb4dcb6f" />
<img width="4160" height="1984" alt="sa_merger_notifications" src="https://github.com/user-attachments/assets/af9fd4ff-d9a0-49cd-98d5-ed05577446aa" />
<img width="1935" height="1343" alt="image" src="https://github.com/user-attachments/assets/c0614390-5aa0-418d-9b54-b096a6baa745" />

# MACROECONOMIC VARIABLES

Explain why we use Nominal GDP (its issues with conflating economic activity and price effects in representing changes in national accounts), how this links to sales and inventory and M&A activity in the economy, how real GDP solves this or the GDP deflator to account for local nexus and market capitalisation, is GNP an alternative and GDP measured in PPP and what is the economic rationale for choosing that? Could we have used other variables? And lastly market capitalisation and the ALSI.

<img width="1484" height="1331" alt="image" src="https://github.com/user-attachments/assets/6b1c76e3-14cc-4a5b-ba36-46b6ed32edee" />
<img width="2385" height="1916" alt="image" src="https://github.com/user-attachments/assets/d7d242ad-a572-4881-8e94-da04c65fa902" />
<img width="4160" height="1984" alt="image" src="https://github.com/user-attachments/assets/3b2cec42-cfb8-4174-b825-186bf29700c7" />

## Why both JSE market capitalisation and the ALSI are retained

Full JSE market capitalisation is the nominal rand value of the whole market and is the variable that most closely matches the Competition Commission's own historical wording ("market capitalisation on the Johannesburg Stock Exchange"). The FTSE/JSE All Share Index (ALSI, J203) is instead a price/capital index covering roughly 99% of eligible Main Board securities by value — an index level, not a rand value — and it is **not a substitute for market capitalisation**: over 2017→2026 the two benchmarks diverge sharply (ALSI +91.0% to a 19 May 2026 snapshot vs +59.9% for full market cap to a 5 June 2026 primary-source snapshot). Both are therefore retained as separate benchmarks: market capitalisation as the primary market-based benchmark (matching the Commission's own language), and the ALSI as a robustness/equity-price comparison. Neither series has a completed full 2026 annual observation; both 2026 points below are dated, point-in-time snapshots, not annual averages.

*Standard workflow reminder, following initial inspection (structure, missingness, summary stats, distribution plots): 1) cleaning decisions, made explicit; 2) feature engineering; 3) bivariate/relationship exploration; 4) formal statistical tests where relevant; 5) model specification and estimation; 6) diagnostics and robustness; 7) communicating results.*

## Construction of the Transaction Sample

**Source.** We evaluate four Bloomberg MA<GO> exports totalling 13,483 rows: two queries, split at 20 August 2015, each run twice with different columns. Matching on type, date, target and acquirer, the ticker exports strictly contain the others, which lose 710 deals and add none; shared values agree throughout. We take the ticker exports as the row universe, 7,105 rows.

Bloomberg data are compiled from filings, press releases, news wires and direct submissions, and unlisted firms are covered subject to disclosure, and coverage "can be thinner for smaller or non-reportable transactions." *Transactions near the intermediate threshold are therefore under-represented, so counts of deals sitting just below the notification boundary are lower bounds.* Sector fields return current, not point-in-time, classifications. *A firm acquired in 2003 carries its 2026 sector, so sector results describe classification today rather than at the deal date.*

**Defects.** Sixteen rows carry a negative turnover across only three distinct values and three firms; we read this as a sign error and correct it. *Left negative, these firms would fail the turnover route automatically and be recorded as asset-driven, so the correction changes which measure is decisive for them.* One deal value is overstated a thousandfold, at forty-eight times the acquirer's balance sheet, which we flag and do not use. *Deal value enters no statutory test, so this affects descriptive statistics on transaction size only.*

The financial columns are firm-level: of 238 acquirers appearing three or more times, every one carries an identical assets figure across all its deals. *These are each firm's latest accounts, not its position at announcement. The median transaction predates the export by fifteen years, so early deals are tested at present-day size against thresholds set in their own era, overstating how many cleared them.* And 11 per cent of targets are asset descriptions — mineral blocks, property portfolios, tower sites. *Missingness is concentrated on the target side by construction: 53 of the 70 three-value records lack a target-side figure, so transferred-firm results rest on a smaller, non-random subset.*

**Selection.** Section 12(1)(b) provides that a merger may be achieved through purchase of shares, an interest, or assets, so we retain deal types M&A, INV and AST. *Excluding asset acquisitions would have narrowed the population below the statutory definition.* We remove 133 repurchases and unbundlings, which record Shareholders as the acquirer and involve no acquisition of control, 106 joint ventures, 15 firms acquiring their own shares, and 166 naming no acquirer or a placeholder. *The repurchases are 28 complete records of large listed firms; retaining them would tilt the sample further towards that group. The placeholders carry no complete records, so that exclusion is immaterial.* We collapse repeated target–acquirer pairs announced within 365 days, keeping the fullest record. *This is the only judgement in the pipeline that materially moves the sample size; a revised bid left uncollapsed would double-weight the same firms.*

**Result.** This leaves 6,511 unique transactions, of which 476 carry three or four financial values: 406 complete and 70 with three, announced between February 1999 and July 2026. Thirty are Withdrawn or Proposed, which Bloomberg defines as rumoured or non-binding, with no definitive agreement signed; all are complete records and all post-date 2009. We retain them, identifiable by status. *They are not a random thirty: retaining them tilts the sample towards recent large listed transactions. Dropping them gives 446.*

**Cleaning funnel**

| Step | Rows |
|---|---:|
| Raw rows across the four exports | 13,483 |
| Ticker exports only | 7,105 |
| After exact and boundary duplicates | 7,095 |
| Deal types M&A, INV and AST | 6,856 |
| Acquirer and target distinct firms | 6,841 |
| Acquirer is a named firm | 6,675 |
| **Unique transactions** | **6,511** |
| &nbsp;&nbsp;with 0 of 4 financial values | 2,362 |
| &nbsp;&nbsp;with 1 of 4 | 398 |
| &nbsp;&nbsp;with 2 of 4 | 3,275 |
| &nbsp;&nbsp;with 3 of 4 | 70 |
| &nbsp;&nbsp;with 4 of 4 | 406 |
| **ANALYSIS SAMPLE (3 or 4 values)** | **476** |
| &nbsp;&nbsp;of which Withdrawn or Proposed | 30 |

*Source: Bloomberg MA<GO>.*

# Data Analysis

```{r load-data, echo=FALSE, message=FALSE}
# I don't want all of the information to be printed out here
library(readr)
library(dplyr)
madata <- read_csv("data/ma_classification_blended.csv", show_col_types = FALSE)
str(madata)
dim(madata)
```

## Adding JSE market capitalisation and the ALSI to the blended data

The blended dataset already carries full JSE market capitalisation, sourced consistently with the dedicated JSE/ALSI workbook (cross-checked below to within 0.04% at every threshold-revision year). The ALSI series is new and is merged in here by year.

```{r merge-alsi, message=FALSE}
library(readxl)
library(tidyr)

jse_file <- "data/JSE_market_cap_and_ALSI_1999_2026.xlsx"

alsi_annual <- read_excel(jse_file, sheet = "ALSI_Annual", skip = 3) %>%
  select(Year, ALSI = `ALSI level`)

mcap_annual <- read_excel(jse_file, sheet = "JSE_Market_Cap", skip = 3) %>%
  select(Year, `Market cap (R million)`)

# 2026 has two snapshot rows in each sheet (dated point-in-time observations,
# not a completed annual figure). Keep the later/primary one in each case:
# ALSI -> 29 May 2026 snapshot; Market cap -> 5 June 2026 JSE-primary snapshot.
alsi_annual <- alsi_annual %>%
  filter(Year != 2026) %>%
  bind_rows(alsi_annual %>% filter(Year == 2026) %>% slice(2))

mcap_annual <- mcap_annual %>%
  filter(Year != 2026) %>%
  bind_rows(mcap_annual %>% filter(Year == 2026) %>% slice(2))

# Cross-check the new file's market cap against what's already in madata
# (should agree to within rounding -- confirms the two sources are consistent)
madata %>%
  distinct(Year, .keep_all = TRUE) %>%
  select(Year, `Market Capitalisation (R million)`) %>%
  inner_join(mcap_annual, by = "Year") %>%
  filter(Year %in% c(2001, 2009, 2017)) %>%
  mutate(diff_pct = round((`Market cap (R million)` - `Market Capitalisation (R million)`) /
                             `Market Capitalisation (R million)` * 100, 3))

# Add ALSI onto madata by Year (market cap already present and verified above)
madata <- madata %>%
  left_join(alsi_annual, by = "Year")
```

## Evolution of the SA threshold (growth by policy change)

```{r threshold-growth-chart}
# ============================================================
# Threshold Growth Chart — Small Multiples, Independent Y-Axes
# ============================================================
library(dplyr)
library(tidyr)
library(ggplot2)

thresholds <- tibble(
  Limb      = c("Intermediate: Combined", "Intermediate: Target",
                "Large: Combined", "Large: Target"),
  base_1999 = c(50, 5, 3500, 100),
  v2001     = c(200, 30, 3500, 100),
  v2009     = c(560, 80, 6600, 190),
  v2017     = c(600, 100, 6600, 190),
  v2026     = c(1000, 200, 9500, 280)
)

long <- thresholds %>%
  pivot_longer(-Limb, names_to = "period_end", values_to = "value") %>%
  group_by(Limb) %>%
  mutate(
    increment  = value - lag(value, default = 0),
    pct_growth = (value / lag(value) - 1) * 100,
    period = factor(period_end,
                     levels = c("base_1999", "v2001", "v2009", "v2017", "v2026"),
                     labels = c("1999 base", "+2001", "+2009", "+2017", "+2026"))
  ) %>%
  ungroup()

ggplot(long, aes(x = 1, y = increment, fill = period)) +
  geom_col(width = 0.6, color = "white") +
  geom_text(
    aes(label = case_when(
      period == "1999 base" ~ paste0("R", value, "m"),
      increment > 0         ~ paste0("+", round(pct_growth), "%"),
      TRUE                  ~ ""
    )),
    position = position_stack(vjust = 0.5),
    color = "white", fontface = "bold", size = 3
  ) +
  facet_wrap(~ Limb, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = c(
    "1999 base" = "#D9D9D9", "+2001" = "#35588A",
    "+2009" = "#A83232", "+2017" = "#4C8C6B", "+2026" = "#B8892B"
  )) +
  labs(
    y = "Threshold Value (R'm)", x = NULL, fill = NULL,
    title = "Growth of Each Threshold Limb Since 1999",
    subtitle = "Each panel independently scaled"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )

ggsave("threshold_growth_small_multiples.png", width = 16, height = 9, dpi = 150)
```

### Grayscale version (for print)

In nominal terms, South Africa's merger notification thresholds have evolved in uneven and episodic steps rather than through smooth or uniform adjustment. The most pronounced increase occurred in the 1999–2001 interval, when the intermediate combined threshold rose by 300 per cent and the intermediate target threshold by 500 per cent, while the large-merger thresholds remained unchanged. A second major adjustment followed in 2001–2009, when all four thresholds increased, with nominal growth again stronger for intermediate mergers than for large mergers. By contrast, the 2009–2017 interval was characterised by very limited adjustment: the intermediate combined threshold increased by only 7.1 per cent, the intermediate target threshold by 25 per cent, and the large thresholds did not change. The 2017–2026 revision marked a renewed upward adjustment, with the intermediate target threshold doubling, the intermediate combined threshold rising by 66.7 per cent, and the large thresholds increasing by 43.9 per cent and 47.4 per cent respectively. Overall, the nominal pattern indicates that threshold growth has been concentrated in a few discrete regulatory revisions, with particularly strong increases at the intermediate-merger boundary.

```{r threshold-growth-grayscale}
grays <- c("1999 base" = "#E8E8E8", "+2001" = "#BFBFBF", "+2009" = "#8C8C8C",
           "+2017" = "#595959", "+2026" = "#262626")

long <- long %>%
  mutate(label_color = if_else(period %in% c("1999 base", "+2001"), "black", "white"))

ggplot(long, aes(x = 1, y = increment, fill = period)) +
  geom_col(width = 0.6, color = "white", linewidth = 0.4) +
  geom_text(
    aes(label = case_when(
          period == "1999 base" ~ paste0("R", value, "m"),
          increment > 0         ~ paste0("+", round(pct_growth), "%"),
          TRUE                  ~ ""),
        color = label_color),
    position = position_stack(vjust = 0.5), fontface = "bold", size = 3, show.legend = FALSE
  ) +
  scale_color_identity() +
  facet_wrap(~ Limb, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = grays) +
  labs(y = "Threshold Value (R'm)", x = NULL, fill = NULL,
       title = "Growth of Each Threshold Limb Since 1999",
       subtitle = "Grayscale — each panel independently scaled") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "top", strip.text = element_text(face = "bold"))

ggsave("threshold_grayscale.png", width = 16, height = 9, dpi = 150)
```

## Research questions and empirical approach

The empirical analysis asks how South Africa's merger notification thresholds have evolved, whether their adjustment can be rationalised by movements in observable measures of the economy, and how South Africa's approach compares with alternative methods of setting merger thresholds. Six benchmarks are now used throughout: CPI, the GDP deflator, nominal GDP, real GDP, full JSE market capitalisation, and the FTSE/JSE ALSI.

### Question 1: What if the Commission had chosen a macroeconomic benchmark in 2001?

Had the Competition Commission chosen in 2001 to anchor the merger notification thresholds to a macroeconomic benchmark, what would each of the four nominal thresholds have been at the subsequent revision points in 2009, 2017 and 2026? For each benchmark, I calculate the percentage growth between 2001 and each subsequent revision date and use this growth to construct the corresponding counterfactual intermediate combined, intermediate target-firm, large combined and large target-firm thresholds, then compare these counterfactual values with the thresholds actually implemented to determine which benchmark most closely tracks the observed statutory path.

I use 2001 rather than 1999 as the principal starting point. The 1999 values were the initial thresholds under the new merger-control regime, while the 2001 revision followed an early review of merger activity. The 1999–2001 adjustment is therefore reported descriptively but is not treated as evidence of a systematic threshold-adjustment rule. Njisane et al. similarly describe the 2001 change as following a review of merger trends since the inception of the Competition Act.

```{r question-1, message=FALSE}
# ============================================================
# QUESTION 1: "Had the commission made an implicit or explicit
# choice to pin the thresholds down to the macroeconomic
# variables we have in our data in 2001 (given that the 1999
# were their initial threshold series, the revision shows
# initial calibration after lessons, so I wouldn't read too
# much into the 1999-2001 revisions) what would each values
# have been at the revision points & what percentage growth
# would they have stipulated, and with commission's current
# data which macro benchmark predicted that better?"
#
# UPDATED: now six benchmarks (CPI, GDP deflator, nominal GDP,
# real GDP, JSE market cap, ALSI) instead of five.
# ============================================================

library(dplyr)
library(tidyr)
library(purrr)
library(kableExtra)

# ---- Macro series: CPI/GDP/Deflator from madata; market cap + ALSI
# from the merged JSE/ALSI series added above ----
macro <- madata %>%
  distinct(Year, .keep_all = TRUE) %>%
  transmute(Year,
            CPI  = `CPI Index (Dec 2024=100)`,
            NGDP = `Nominal GDP (R million)`,
            RGDP = `Real GDP (R million, 2015 prices)`,
            DEFL = `GDP Deflator (Index, 2015=100)`,
            MCAP = `Market Capitalisation (R million)`,
            ALSI = ALSI) %>%
  arrange(Year) %>%
  filter(!is.na(CPI))  # drops the incomplete 2026 row in madata itself

# ---- 2026-so-far values for the GDP-derived series (partial-year, as of
# Sept 2026 -- CPI: Jan-Jul avg; Real/Nominal GDP: Q1-Q2/Q1 annualised).
# Market cap and ALSI 2026 already came from the dated JSE/ALSI snapshots
# merged in above, so only the GDP-derived figures need overriding here. ----
# ---- 2026-so-far values (partial-year, as of Sept 2026).
# CPI: Jan-Jul avg; Real/Nominal GDP: Q1-Q2/Q1 annualised (GDP-derived,
# entered here). Market cap and ALSI 2026 come from the dated JSE/ALSI
# snapshots already sitting in mcap_annual / alsi_annual from the merge
# step above -- pulled in here rather than hardcoded a second time. ----

row_2026 <- tibble(
  Year = 2026,
  CPI  = 105.8,
  NGDP = 8025000,
  RGDP = 4720000,
  DEFL = 8025000 / 4720000 * 100,
  MCAP = mcap_annual %>% filter(Year == 2026) %>% pull(`Market cap (R million)`),
  ALSI = alsi_annual %>% filter(Year == 2026) %>% pull(ALSI)
)

macro <- macro %>%
  filter(Year != 2026) %>%   # drop any incomplete/partial 2026 row first
  bind_rows(row_2026) %>%
  arrange(Year)

# sanity check
macro %>% filter(Year == 2026)
macro <- macro %>%
  rows_update(
    tibble(Year = 2026, CPI = 105.8, NGDP = 8025000, RGDP = 4720000,
           DEFL = 8025000/4720000*100),
    by = "Year"
  )

# ---- Threshold data (Table 2, legislated figures) ----
actual <- tribble(
  ~Year, ~IC,   ~IT,  ~LC,   ~LT,
  1999,   50,    5,   3500,  100,
  2001,   200,   30,  3500,  100,
  2009,   560,   80,  6600,  190,
  2017,   600,   100, 6600,  190,
  2026,   1000,  200, 9500,  280
)
limb_names  <- c(IC = "Interm: Combined", IT = "Interm: Target",
                  LC = "Large: Combined",  LT = "Large: Target")
bench_names <- c(CPI = "CPI", NGDP = "Nominal GDP", RGDP = "Real GDP",
                  DEFL = "GDP Deflator", MCAP = "Market Cap", ALSI = "ALSI")
base_year <- 2001

predict_years <- c(2009, 2017, 2026)
base_macro  <- macro %>% filter(Year == base_year)
base_thresh <- actual %>% filter(Year == base_year)

q1_data <- map_dfr(predict_years, function(yr) {
  yr_macro <- macro %>% filter(Year == yr)
  map_dfr(names(limb_names), function(code) {
    actual_val <- actual %>% filter(Year == yr) %>% pull(all_of(code))
    base_val   <- base_thresh %>% pull(all_of(code))
    map_dfr(names(bench_names), function(b) {
      gf <- yr_macro[[b]] / base_macro[[b]]
      predicted <- base_val * gf
      tibble(Year = yr, Limb = limb_names[[code]], Benchmark = bench_names[[b]],
             `Predicted (R'm)` = round(predicted, 1),
             `Predicted Growth %` = round((gf - 1) * 100, 1),
             `Error vs Actual %` = round((predicted - actual_val) / actual_val * 100, 1))
    })
  })
})

# ---- ONE INVERTED (metrics-as-rows) TABLE PER YEAR, sized to fit a page ----
for (yr in predict_years) {
  yr_block <- q1_data %>% filter(Year == yr)
  wide <- yr_block %>%
    pivot_longer(c(`Predicted (R'm)`, `Predicted Growth %`, `Error vs Actual %`),
                 names_to = "Metric", values_to = "value") %>%
    unite("row_label", Benchmark, Metric, sep = " \u2014 ") %>%
    pivot_wider(id_cols = row_label, names_from = Limb, values_from = value)

  actual_row <- actual %>% filter(Year == yr) %>%
    pivot_longer(-Year, names_to = "code", values_to = "val") %>%
    mutate(Limb = limb_names[code]) %>%
    select(Limb, val) %>%
    pivot_wider(names_from = Limb, values_from = val) %>%
    mutate(row_label = "Actual Value (R'm)", .before = 1)

  final_tbl <- bind_rows(actual_row, wide)

  print(
    kbl(final_tbl, format = "latex", booktabs = TRUE, longtable = FALSE,
        col.names = c("", names(final_tbl)[-1]),
        caption = paste0("Threshold values predicted under each macro benchmark vs actual, ", yr,
                          " (single base = 2001)"),
        label = paste0("tab:q1-", yr)) %>%
      kable_styling(latex_options = c("scale_down", "hold_position"),
                    font_size = 8) %>%
      column_spec(1, width = "5cm") %>%
      row_spec(0, bold = TRUE)
  )
}
```

### Question 2: What if the benchmark were reconsidered at each revision?

Had the Commission reconsidered the appropriate benchmark at each threshold revision, what adjustment would each macroeconomic benchmark have implied over the individual intervals 2001–2009, 2009–2017 and 2017–2026? Unlike Question 1, the thresholds are re-anchored to the actual statutory value at the beginning of each interval. This allows the analysis to ask whether different revisions appear to have followed different economic benchmarks rather than assuming a single adjustment rule throughout the entire period.

This is particularly relevant because the Commission's stated rationale has differed across revisions. Njisane et al. report that the 2009 adjustment was informed by nominal GDP and market capitalisation, whereas the 2017 adjustment was based on real GDP growth.

```{r question-2, message=FALSE}
# ============================================================
# QUESTION 2: "Had this choice been made within interval, that
# is reanchoring to 2001, 2009, 2017, 2026, what would have
# this implied, given the commission's reports states different
# reasons at each interval? Just verify the macro benchmark
# data because some of them are already indexed to 2024 etc
# (handle that)."
#
# Note: base-year indexing is handled automatically here because
# every comparison uses a ratio of the index at two points in
# time -- the arbitrary base year (Dec 2024 for CPI, 2015 for the
# GDP Deflator) cancels out algebraically in a ratio.
# 1999-2001 is EXCLUDED: it's the initial post-enactment
# calibration, not a benchmark-driven revision (see Question 1
# framing above and the negative-control check in the appendix).
#
# UPDATED: now six benchmarks, ALSI added.
# ============================================================

intervals <- list(c(2001, 2009), c(2009, 2017), c(2017, 2026))

q2_data <- map_dfr(intervals, function(iv) {
  start_yr <- iv[1]; end_yr <- iv[2]
  start_macro <- macro %>% filter(Year == start_yr)
  end_macro   <- macro %>% filter(Year == end_yr)
  start_thresh <- actual %>% filter(Year == start_yr)
  end_thresh   <- actual %>% filter(Year == end_yr)

  map_dfr(names(limb_names), function(code) {
    a_start <- start_thresh %>% pull(all_of(code))
    a_end   <- end_thresh %>% pull(all_of(code))
    map_dfr(names(bench_names), function(b) {
      gf <- end_macro[[b]] / start_macro[[b]]
      implied <- a_start * gf
      tibble(Interval = paste0(start_yr, "\u2192", end_yr), Limb = limb_names[[code]],
             Benchmark = bench_names[[b]],
             `Implied (R'm)` = round(implied, 1),
             `Implied Growth %` = round((gf - 1) * 100, 1),
             `Error vs Actual %` = round((implied - a_end) / a_end * 100, 1))
    })
  })
})

# ---- ONE INVERTED TABLE PER INTERVAL ----
for (iv in intervals) {
  label <- paste0(iv[1], "\u2192", iv[2])
  iv_block <- q2_data %>% filter(Interval == label)
  wide <- iv_block %>%
    pivot_longer(c(`Implied (R'm)`, `Implied Growth %`, `Error vs Actual %`),
                 names_to = "Metric", values_to = "value") %>%
    unite("row_label", Benchmark, Metric, sep = " \u2014 ") %>%
    pivot_wider(id_cols = row_label, names_from = Limb, values_from = value)

  actual_rows <- bind_rows(
    actual %>% filter(Year == iv[1]) %>%
      pivot_longer(-Year, names_to="code", values_to="val") %>%
      mutate(Limb=limb_names[code]) %>% select(Limb,val) %>%
      pivot_wider(names_from=Limb, values_from=val) %>%
      mutate(row_label = paste0("Actual Start (", iv[1], ", R'm)"), .before=1),
    actual %>% filter(Year == iv[2]) %>%
      pivot_longer(-Year, names_to="code", values_to="val") %>%
      mutate(Limb=limb_names[code]) %>% select(Limb,val) %>%
      pivot_wider(names_from=Limb, values_from=val) %>%
      mutate(row_label = paste0("Actual End (", iv[2], ", R'm)"), .before=1)
  )

  final_tbl <- bind_rows(actual_rows, wide)

  print(
    kbl(final_tbl, format = "latex", booktabs = TRUE,
        col.names = c("", names(final_tbl)[-1]),
        caption = paste0("Threshold growth reanchored per interval, ", label,
                          " \u2014 which macro benchmark implied best fit (six benchmarks)"),
        label = paste0("tab:q2-", gsub("\u2192","-",label))) %>%
      kable_styling(latex_options = c("scale_down", "hold_position"),
                    font_size = 8) %>%
      column_spec(1, width = "5cm") %>%
      row_spec(0, bold = TRUE)
  )
}
```

**Updated best-fit summary (six benchmarks).** Adding the ALSI changes two results relative to the five-benchmark version: ALSI is now the closest-fitting benchmark for both Intermediate limbs over 2001→2009 (errors of −5.4% and −0.6%, ahead of Nominal GDP's −14.4%/−10.1%), and for Intermediate: Target over 2017→2026 (−3.7%, ahead of Market Cap's −20.0% on the updated, primary-sourced market-cap figure). Real GDP remains the unique best fit across all four limbs for 2009→2017, and CPI remains the best fit for both Large limbs over 2017→2026 (errors of 2.9% and 0.5%) — those two findings are unchanged by adding the ALSI.

### Question 3: How large were the revisions relative to alternative threshold-setting methods?

How do the actual statutory adjustments compare with the threshold increases examined by Njisane et al. and with the World Bank's economy-size benchmarking approach?

Njisane et al.'s scenarios are used as sensitivity benchmarks. Their target-firm thresholds are increased by 5%, 10%, 15% and 20%, while combined thresholds are increased by 10%, 20%, 30%, 40% and 45%. They also examine joint increases of 15% in the target threshold with either 20% or 40% in the combined threshold. These percentage bands are compared with the actual percentage change in each South African threshold over the 2001–2009, 2009–2017 and 2017–2026 intervals.

As a sensitivity exercise, the percentage-adjustment scenarios considered by Njisane et al. are applied to the statutory threshold in force at the beginning of each revision interval. This extends their scenario-based approach, originally applied to the October 2017 thresholds and the 2016–2018 merger database, to the historical threshold path. How does the magnitude of the actual revision compare with moderate threshold shocks that prior South African research considered plausible enough to test for error-cost consequences?

```{r question-3-njisane, message=FALSE}
library(tidyverse)
library(kableExtra)

# ============================================================
# QUESTION 3 (Njisane et al. 2021 sensitivity comparison)
# How large were SA's actual threshold revisions relative to
# the target/combined increases Njisane et al. tested, and what
# would their two selected joint scenarios have implied at each
# SA revision point?
# ============================================================

# ---- 1. Actual SA threshold history (R million) ----
sa_thresholds <- tribble(
  ~year, ~int_combined, ~int_target, ~large_combined, ~large_target,
  2001,      200,           30,           3500,          100,
  2009,      560,           80,           6600,          190,
  2017,      600,          100,           6600,          190,
  2026,     1000,          200,           9500,          280
)

actual_growth <- sa_thresholds %>%
  pivot_longer(-year, names_to = "limb_code", values_to = "threshold") %>%
  group_by(limb_code) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    start_year        = lag(year),
    start_threshold   = lag(threshold),
    interval          = paste0(start_year, "\u2192", year),
    actual_growth_pct = 100 * (threshold / start_threshold - 1)
  ) %>%
  ungroup() %>%
  filter(!is.na(start_year)) %>%
  mutate(
    limb_type = if_else(str_detect(limb_code, "target"), "Target", "Combined"),
    limb = recode(limb_code,
                  int_combined   = "Intermediate: Combined",
                  int_target     = "Intermediate: Target",
                  large_combined = "Large: Combined",
                  large_target   = "Large: Target")
  ) %>%
  select(interval, limb, limb_type,
         start_threshold, actual_end_threshold = threshold, actual_growth_pct)

# ---- 2. Njisane et al. (2021) single-limb sensitivity scenarios ----
njisane_scenarios <- tribble(
  ~limb_type, ~scenario_pct,
  "Target",    5,  "Target",   10, "Target",   15, "Target",   20,
  "Combined", 10,  "Combined", 20, "Combined", 30, "Combined", 40, "Combined", 45
)

njisane_range <- njisane_scenarios %>%
  group_by(limb_type) %>%
  summarise(njisane_min = min(scenario_pct),
            njisane_max = max(scenario_pct), .groups = "drop")

closest_njisane <- actual_growth %>%
  left_join(njisane_scenarios, by = "limb_type") %>%
  mutate(diff_pp = actual_growth_pct - scenario_pct) %>%
  group_by(interval, limb) %>%
  slice_min(abs(diff_pp), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(interval, limb, closest_njisane_pct = scenario_pct, closest_diff_pp = diff_pp)

# ---- 3. Njisane's two selected JOINT scenarios ----
joint_scenarios <- tribble(
  ~scenario_label,             ~target_pct, ~combined_pct,
  "15%T / 20%C implied",       15,          20,
  "15%T / 40%C implied",       15,          40
)

njisane_joint_wide <- actual_growth %>%
  select(interval, limb, limb_type, start_threshold) %>%
  crossing(joint_scenarios) %>%
  mutate(
    applied_pct       = if_else(limb_type == "Target", target_pct, combined_pct),
    implied_threshold = start_threshold * (1 + applied_pct / 100)
  ) %>%
  select(interval, limb, scenario_label, implied_threshold) %>%
  pivot_wider(names_from = scenario_label, values_from = implied_threshold)

# ---- 4. ONE combined table ----
q3_table <- actual_growth %>%
  left_join(njisane_range,      by = "limb_type") %>%
  mutate(
    njisane_range_label = paste0(njisane_min, "\u2013", njisane_max, "%"),
    position_vs_range = case_when(
      actual_growth_pct < njisane_min ~ "Below range",
      actual_growth_pct > njisane_max ~ "Above range",
      TRUE ~ "Within range"
    )
  ) %>%
  left_join(closest_njisane,    by = c("interval", "limb")) %>%
  left_join(njisane_joint_wide, by = c("interval", "limb")) %>%
  transmute(
    Interval              = interval,
    Limb                  = limb,
    `Start (R'm)`         = round(start_threshold, 1),
    `Actual End (R'm)`    = round(actual_end_threshold, 1),
    `Actual Growth %`     = round(actual_growth_pct, 1),
    `Njisane Range`       = njisane_range_label,
    `Position`            = position_vs_range,
    `Closest Njisane %`   = closest_njisane_pct,
    `Diff (pp)`           = round(closest_diff_pp, 1),
    `15%T/20%C (R'm)`     = round(`15%T / 20%C implied`, 1),
    `15%T/40%C (R'm)`     = round(`15%T / 40%C implied`, 1)
  ) %>%
  arrange(Interval, Limb)

print(
  kbl(q3_table, format = "latex", booktabs = TRUE, longtable = TRUE,
      caption = "SA threshold revisions vs. Njisane et al. (2021) sensitivity scenarios",
      label = "tab:q3-njisane") %>%
    kable_styling(latex_options = c("scale_down", "hold_position", "repeat_header"),
                  font_size = 7) %>%
    row_spec(0, bold = TRUE)
)



# ---- 5. ONE graph: actual growth vs Njisane's tested range, faceted by interval ----
plot_data <- actual_growth %>% left_join(njisane_range, by = "limb_type")

ggplot(plot_data, aes(x = limb, y = actual_growth_pct)) +
  geom_col(fill = "grey40", width = 0.6) +
  geom_errorbar(aes(ymin = njisane_min, ymax = njisane_max),
                width = 0.15, linewidth = 0.7, colour = "black") +
  facet_wrap(~ interval, nrow = 1, scales = "free_x") +
  labs(
    title = "Actual SA threshold growth vs. Njisane et al. (2021) tested range",
    subtitle = "Bars: actual nominal growth. Error bars: min\u2013max of Njisane's tested scenarios for that limb type.",
    x = NULL, y = "Nominal threshold growth (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Macroeconomic data and measurement

All benchmark comparisons are based on growth rates rather than the absolute levels of the macroeconomic indices. Rebasings therefore do not affect the results provided that each growth calculation uses a consistently linked historical series — this applies equally to the CPI (Dec 2024 = 100), the GDP deflator (2015 = 100), and the ALSI (a price index with no fixed base year in this compilation), since every comparison below uses a ratio of the index at two points in time and the arbitrary base cancels out algebraically.

CPI is measured using the official headline CPI series. The GDP deflator is treated as a price-level measure and is constructed or obtained from a consistent national-accounts vintage. Nominal GDP is measured at current prices, while real GDP is measured at constant 2015 prices. Full JSE market capitalisation is the nominal rand value of the whole market; the ALSI is the FTSE/JSE All Share Index (J203), an equity price/capital index rather than a rand value, retained as a distinct robustness benchmark for the reasons set out above.

For each benchmark \(B\), the counterfactual threshold is calculated as:

\[
T_{t}^{B}=T_{s}\left(\frac{B_t}{B_s}\right),
\]

where \(T_s\) is the statutory threshold at the chosen anchor date, \(B_s\) is the benchmark value at that date, and \(B_t\) is its value at the subsequent revision date. The implied percentage adjustment is:

\[
g_{s,t}^{B}=\left(\frac{B_t}{B_s}-1\right)\times100.
\]

For Question 1, \(s=2001\) throughout. For Question 2, \(s\) is reset to the statutory threshold at the beginning of each interval: 2001, 2009 and 2017.



```{r}
# ============================================================
# QUESTION 4: How have the comparator jurisdictions evolved?
#
# "What were the initial and current merger-notification
# thresholds in the United States, European Union and Brazil;
# how and why have these thresholds changed; what determines
# their adjustment where an adjustment mechanism exists; and
# how does the design of each notification regime differ from
# South Africa's?"
#
# World Bank cross-country regression is deliberately excluded
# from this table -- kept as a separate exercise, since the
# threshold architectures here are not all comparable on one
# regression (transaction-value vs worldwide vs domestic
# turnover), per the note under "The World Bank log graph".
#
# Facts below verified independently against primary/authoritative
# sources (FTC press releases for the US; CADE/Lexology/Law
# 12,529 sources for Brazil) before entering the dissertation.
# ============================================================

library(tidyverse)
library(kableExtra)

comparators <- tribble(
  ~Jurisdiction, ~`Starting Point`, ~`Current (2026)`, ~Metric, ~`Adjustment Mechanism`, ~`Economic Anchor`, ~Frequency,

  "United States",
  "2001: transaction > \\$50m (reset from original 1976 \\$15m floor); size-of-person \\$10m/\\$100m; transaction > \\$200m regardless of size-of-person",
  "Transaction > \\$133.9m; size-of-person \\$26.8m/\\$267.8m; > \\$535.5m irrespective of size-of-person",
  "Transaction value + party assets/sales",
  "Automatic statutory indexation",
  "GNP",
  "Annual, from FY2005",

  "European Union",
  "1990: ECU 5bn worldwide combined + ECU 250m Community turnover for each of at least two firms",
  "\\euro5bn + \\euro250m (main route, unchanged); second (lower, multi-Member-State) route added 1 March 1998",
  "Turnover (worldwide + Community/Member-State)",
  "No automatic indexation; legislative review/revision only",
  "None mechanically specified",
  "Irregular / legislative",

  "Brazil",
  "2011 statute (Law 12,529/2011): R\\$400m + R\\$30m local group turnover",
  "R\\$750m + R\\$75m (set by Interministerial Ordinance 994/2012, unchanged since; a rise to R\\$1bn/R\\$200m was under discussion as of Jan 2025 but not confirmed enacted)",
  "Local turnover of two economic groups",
  "Discretionary: CADE recommendation + interministerial ordinance",
  "No automatic macroeconomic formula",
  "Irregular; unchanged since 2012",

  "South Africa",
  "1999: Intermediate R50m + R5m; Large R3.5bn + R100m",
  "Intermediate R1bn + R200m; Large R9.5bn + R280m",
  "Local turnover OR assets; combined + target limbs",
  "Discretionary Minister/Commission determination",
  "Has varied between revisions (see Q1/Q2)",
  "Episodic"
)

print(
  kbl(comparators, format = "latex", booktabs = TRUE, longtable = TRUE, escape = FALSE,
      caption = "Comparator merger-notification threshold regimes: United States, European Union, Brazil and South Africa",
      label = "tab:q4-comparators") %>%
    kable_styling(latex_options = c("scale_down", "hold_position", "repeat_header"),
                  font_size = 7) %>%
    column_spec(1, bold = TRUE, width = "2cm") %>%
    column_spec(2, width = "3.5cm") %>%
    column_spec(3, width = "3.5cm") %>%
    column_spec(4, width = "2.5cm") %>%
    column_spec(5, width = "2.5cm") %>%
    column_spec(6, width = "2cm") %>%
    column_spec(7, width = "1.8cm") %>%
    row_spec(0, bold = TRUE)
)
```

# 4 Descriptive Analysis

The threshold data reveal substantial variation in the evolution of South Africa's merger-notification regime. The size and timing of threshold adjustments differ across the four statutory values, and the economic benchmark that most closely reproduces these adjustments also varies across revision periods. In particular, there is little descriptive evidence that South Africa has followed a single, stable indexation rule over the life of the regime. The analysis below first considers the statutory evolution of the thresholds and the experience of the three comparator jurisdictions. It then compares the South African thresholds with alternative economic benchmark paths.

## 4.1 The Evolution of South African Merger Thresholds

Figure 1 illustrates the evolution of South Africa's intermediate and large merger thresholds since the introduction of the current merger-control regime in 1999. The graph confirms that threshold adjustment has been intermittent and that the four statutory values have not moved proportionately.

**[Figure 1: South African merger-notification thresholds, 1999–2026]**

The initial intermediate thresholds were R50 million for the combined-firm requirement and R5 million for the target-firm requirement. The corresponding large-merger thresholds were R3.5 billion and R100 million. The first revision, in 2001, increased the intermediate thresholds to R200 million and R30 million respectively but left both large-merger thresholds unchanged. Njisane et al. describe this first adjustment as following a review of merger trends since the introduction of the regime. 

The next revision, implemented in 2009, affected all four values. The intermediate combined threshold increased from R200 million to R560 million and the target threshold from R30 million to R80 million. The large combined threshold increased from R3.5 billion to R6.6 billion, while the large target threshold increased from R100 million to R190 million. These represent increases of 180.0%, 166.7%, 88.6% and 90.0%, respectively.

Threshold adjustment was considerably more limited between 2009 and 2017. The intermediate combined threshold increased by 7.1%, from R560 million to R600 million, and the target threshold increased by 25%, from R80 million to R100 million. Neither large-merger threshold changed. Njisane et al. report that the 2009 revision was informed by nominal GDP and market capitalisation, while the subsequent revision was based on changes in real GDP. 

The 2026 revision represents a comparatively large adjustment, particularly at the intermediate-merger boundary. The intermediate combined threshold increased from R600 million to R1 billion, an increase of 66.7%, while the target threshold doubled from R100 million to R200 million. The large combined threshold increased from R6.6 billion to R9.5 billion, or 43.9%, while the large target threshold increased by 47.4%, from R190 million to R280 million.

Figure 1 therefore highlights two features of the South African threshold history. First, long periods in which nominal values remained unchanged or changed only slightly have been followed by relatively large discrete adjustments. Second, the adjustments have differed across the intermediate and large thresholds. The threshold regime consequently cannot be characterised accurately as a single monetary boundary that has simply been moved upwards periodically.

This distinction is relevant to the historical rationale for threshold revision. Before the 2009 adjustment, the Commission argued that the unchanged thresholds had declined in economic significance as prices, nominal output and JSE market capitalisation increased. The Commission therefore viewed the fixed nominal values as widening the regulatory net over time. This rationale is consistent with the more general threshold-setting problem identified by the ICN: fixed notification thresholds may require periodic review as economic conditions change, although the ICN does not prescribe a single index that should govern this adjustment. 

### 4.1.1 Comparison with Selected Jurisdictions

Table 1 compares the structure and adjustment mechanism of the South African regime with those of the United States, European Union and Brazil.

**[Table 1: Merger-notification threshold design and adjustment across selected jurisdictions]**

The comparison shows substantial differences in both the construction and maintenance of merger thresholds. The United States provides the clearest example of a systematic adjustment rule. Its principal HSR monetary thresholds are adjusted annually with changes in gross national product. The adjustment mechanism is therefore determined in advance rather than selected periodically.

The European Union follows a different approach. Its principal merger-control thresholds are based on combined worldwide turnover and the individual EU turnover of the merging parties. The principal monetary values have remained broadly unchanged, while jurisdictional coverage is supplemented by an additional multi-Member-State route and referral mechanisms. There is no automatic CPI-, GDP- or GNP-based adjustment.

Brazil also relies principally on turnover, but uses separate domestic turnover requirements for two economic groups rather than a South African-style combined-firm and target-firm test. Its current monetary thresholds were established in 2012 and are not subject to automatic macroeconomic indexation.

South Africa therefore differs from all three comparators. It uses separate intermediate and large classifications and, within each category, requires both a combined-firm requirement and a target-firm requirement. It also permits assets or turnover to determine whether the monetary screen is satisfied. More broadly, the comparison confirms that there is no common international approach to maintaining merger thresholds over time. This is consistent with the ICN's survey evidence, which identifies historical benchmarking, comparisons with similarly situated jurisdictions and inflation adjustment as alternative approaches used by competition authorities. 

The international comparison is therefore informative primarily about **adjustment architecture**, rather than about the relative nominal level of thresholds. The US represents automatic indexation; the EU largely maintains long-lived nominal turnover criteria; Brazil allows discretionary adjustment; and South Africa has relied on periodic revisions motivated by different economic considerations.

---

## 4.2 South African Thresholds and Economic Benchmarks

We next consider whether the statutory adjustments resemble changes in broader measures of economic activity. Figure 2 compares the evolution of the actual thresholds with counterfactual paths based on CPI, the GDP deflator, nominal GDP, real GDP, JSE market capitalisation and the FTSE/JSE All Share Index.

**[Figure 2: Actual and benchmark-implied merger-threshold paths]**

The benchmark paths should not be interpreted as alternative estimates of a uniquely correct threshold. Rather, they illustrate different ways in which the economic significance of a fixed monetary threshold might be maintained. This distinction is particularly relevant to South Africa because the Commission itself has referred to different economic measures when motivating previous revisions. 

For purposes of comparison, a benchmark is said to track an actual threshold more closely when the value implied by growth in that benchmark has a smaller absolute percentage deviation from the statutory threshold eventually adopted. Tracking therefore refers only to **descriptive fit**; it does not establish that the Commission used that variable in determining the threshold.

### 4.2.1 The 2001–2009 Revision

Figure 3(a) presents the benchmark-implied changes between 2001 and 2009. The results show a clear difference between the intermediate and large-merger thresholds.

**[Figure 3: Actual and benchmark-implied threshold adjustments by revision interval]**

The intermediate combined threshold increased by 180%, while the intermediate target threshold increased by 166.7%. Among the benchmark paths, the ALSI provides the closest fit to both intermediate thresholds. The ALSI-implied combined threshold is 5.4% below the R560 million statutory value, while the corresponding target-firm threshold is only 0.6% below the R80 million statutory value.

The result differs for large mergers. The actual large combined and target thresholds increased by 88.6% and 90%, respectively. The GDP deflator provides the closest fit to both, producing counterfactual thresholds 4.3% and 5.1% below the actual values.

Thus, the 2009 revision does not appear to correspond to a common adjustment factor across the four thresholds. The intermediate thresholds most closely resemble growth in the equity-price benchmark, whereas the large-merger thresholds more closely resemble the GDP deflator.

This is noteworthy given the Commission's stated basis for the revision. Njisane et al. report that the 2009 thresholds were adjusted with reference to nominal GDP and market capitalisation.  The descriptive results show that these variables help explain the economic context in which the revision occurred, but neither reproduces all four statutory outcomes most closely.

### 4.2.2 The 2009–2017 Revision

The pattern changes considerably over the following revision period.

The intermediate combined threshold increased by only 7.1%, while the target threshold increased by 25%. Both large thresholds remained unchanged. Real GDP provides the closest of the six benchmark paths across all four statutory values.

For the intermediate combined threshold, the real-GDP-implied value is 8.9% above the R600 million statutory threshold. For the target threshold, the counterfactual lies 6.6% below the actual R100 million value. The real-GDP counterfactual exceeds both unchanged large thresholds by 16.7%.

The result is broadly consistent with the institutional account of the revision. Njisane et al. report that the adjustment implemented in 2017 was based on changes in real GDP from 2009 to 2016.  Nevertheless, Figure 3(b) also makes clear that even the closest benchmark does not reproduce the four statutory values exactly. In particular, the absence of any adjustment to the large thresholds implies that those values declined relative to every positive-growth benchmark considered.

### 4.2.3 The 2017–2026-to-Date Revision

Figure 3(c) reports the corresponding results for the latest revision period. The endpoint is deliberately measured on a **2026-to-date basis**, using the latest available observations described in the data section.

The intermediate combined threshold increased by 66.7%. JSE market capitalisation provides the closest benchmark-implied value, lying approximately 4.1% below the actual R1 billion threshold. The intermediate target threshold increased by 100%. Here the ALSI provides the closest fit, with the implied value approximately 3.7% below the actual R200 million threshold.

The large thresholds again display a different pattern. The actual combined threshold increased by 43.9% and the target threshold by 47.4%. CPI provides the closest fit to both. The CPI-implied large combined threshold is approximately 2.9% above the actual R9.5 billion value, while the implied target threshold is approximately 0.5% above the actual R280 million threshold.

Table 2 summarises these results.

**Table 2. Closest benchmark by threshold and revision interval**

| Revision interval | Intermediate combined             | Intermediate target | Large combined       | Large target         |
| ----------------- | --------------------------------- | ------------------- | -------------------- | -------------------- |
| 2001–2009         | ALSI (-5.4%)                      | ALSI (-0.6%)        | GDP deflator (-4.3%) | GDP deflator (-5.1%) |
| 2009–2017         | Real GDP (+8.9%)                  | Real GDP (-6.6%)    | Real GDP (+16.7%)    | Real GDP (+16.7%)    |
| 2017–2026-to-date | JSE market capitalisation (-4.1%) | ALSI (-3.7%)        | CPI (+2.9%)          | CPI (+0.5%)          |

*Notes: Values in parentheses report the percentage deviation of the benchmark-implied value from the actual statutory threshold. A positive value indicates that the benchmark-implied threshold exceeds the actual threshold.*

The main result from Table 2 is that there is **no economic benchmark that consistently provides the closest fit across periods and threshold limbs**. Real GDP provides the most uniform fit in the 2009–2017 period, but this pattern does not extend backwards or forwards. The results therefore provide little descriptive support for an implicit stable indexation rule governing South African merger thresholds.

This finding does not imply that the economic indicators cited by the Commission were irrelevant to the adjustment process. Rather, the resulting statutory values appear to reflect a more differentiated adjustment than would follow from mechanically applying any one of the benchmark series.

---

## 4.3 Actual and Counterfactual Threshold Paths

The preceding analysis re-anchors the counterfactual calculation at the beginning of each statutory revision interval. We now consider a stricter counterfactual: what would the threshold paths have looked like had South Africa selected a particular economic benchmark in 2001 and applied it consistently thereafter?

Figure 4 plots these cumulative benchmark paths against the actual thresholds.

**[Figure 4: Actual thresholds and 2001-anchored counterfactual paths]**

The figure highlights substantial divergence between the actual statutory path and several of the mechanical adjustment rules. Importantly, the direction of this divergence is not the same for the intermediate and large-merger thresholds.

### 4.3.1 Intermediate Thresholds

For the intermediate thresholds, the statutory path initially rises sharply. The 2009 values are close to the ALSI-implied path, as noted above. Thereafter, however, the actual and benchmark paths diverge substantially.

By 2017, several counterfactual rules imply threshold values that differ materially from the statutory values. The limited adjustment of the intermediate combined threshold between 2009 and 2017 causes it to fall relative to most measures of nominal economic activity and prices. The intermediate target threshold shows less divergence because its 25% statutory increase over the period was larger.

The 2026 revision reverses much of this pattern. The combined threshold rises by 66.7%, while the target threshold doubles. Consequently, the intermediate thresholds move upwards relative to several benchmark-maintenance paths. The target threshold shows the strongest movement: its increase is considerably larger than that implied by price-based benchmarks and is closest to the growth in the ALSI over the latest interval.

The result is therefore not a monotonic widening or narrowing of the notification boundary. Rather, the relative position of the statutory threshold varies substantially across regimes.

### 4.3.2 Large-Merger Thresholds

The large thresholds show a different cumulative pattern.

The 2001 values remained unchanged from the initial 1999 regime. Although both were increased substantially in 2009, they were then held constant through 2017. As a result, benchmark paths linked to prices, nominal GDP and financial-market measures increasingly exceed the statutory large-merger thresholds over this period.

The 2026 revision reduces some of this divergence. In particular, the latest increases in the two large thresholds are close to the increase in CPI over the 2017–2026-to-date interval. The cumulative 2001-anchored paths nevertheless show that this latest adjustment does not undo the consequences of earlier periods in which the statutory thresholds grew considerably more slowly than some of the economic benchmarks.

Figures 4(a) and 4(b) therefore reinforce the finding that the intermediate and large thresholds have followed distinct trajectories. A single index applied uniformly to the four 2001 starting values would preserve the ratios between those values through time. The actual South African regime does not do so. The relative structure of the threshold system has itself changed.

### 4.3.3 Interpretation of the Counterfactual Gaps

The gaps between actual and benchmark-implied thresholds have a direct screening interpretation. Where an actual threshold lies above its counterfactual value, the statutory notification boundary is higher than it would have been under that benchmark. Conversely, where the actual value lies below the counterfactual, the statutory boundary is lower.

This distinction is consistent with the error-cost framework used in the threshold literature. The ICN characterises notification-threshold design as balancing the private and administrative costs of capturing transactions that are unlikely to raise competition concerns against the cost of allowing potentially problematic transactions to escape mandatory review.  Njisane et al. apply a similar framework to South Africa, describing these as Type I and Type II error costs. 

The descriptive results suggest that the relative breadth of the South African notification screen has changed over time. The limited statutory adjustment between 2009 and 2017 caused the thresholds to decline relative to most nominal and price benchmarks. In this relative sense, the notification screen became broader. The 2026 revision moves in the opposite direction at the intermediate boundary, where the statutory increases exceed several of the benchmark-maintenance paths.

The result should be interpreted cautiously. The counterfactual gaps do not identify policymakers' preferences. A lower statutory threshold relative to an economic benchmark is consistent with maintaining broader mandatory coverage, while a higher relative threshold is consistent with a narrower screen. Neither establishes why policymakers chose the observed values.

This distinction is particularly important because objective financial thresholds are imperfect screens for substantive competition concerns. The ICN notes that turnover- and asset-based thresholds necessarily capture transactions that ultimately raise no competition concern while potentially excluding others that may be competitively significant.  The threshold comparison therefore identifies changes in **screening coverage**, not changes in the underlying competitive harm of mergers.

The transaction-level analysis that follows considers whether these differences in nominal and relative threshold levels are economically consequential. Specifically, it examines whether application of the alternative threshold paths materially changes the distribution of observed transactions between small, intermediate and large merger classifications.



 

